\texttt{dispRity} demo for palaeobiologists

### 2016-03-17 doc/dispRity-palaeo-demo.Rnw
Minor typo in loading strap (library(strap))
All manual updated

This demo aims to give quick overview of the \dispRity package (v.\Sexpr{packageVersion("dispRity")}) for palaeobiology analysis.
Please refer to \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} for other vignettes, namely the \dispRity tutorial that explains the functions in more detail.

To keep it short, this package allows to easily perform \textbf{disparity-through-time} analysis.
This type of analysis often starts by ordinating \textbf{morphometric} or \textbf{cladistic} data into a multidimensional object hereafter called the \textbf{morphospace}.
One might then be interested how the occupancy of the morphospace changes through time by measuring a summary metric of this morphospace through time, called \textbf{disparity}.
In the morphospace case, disparity might be seen as a value summarising the diversity of morphologies.

This demo showcases a typical disparity-through-time analyse: we are going to test whether the disparity changed through time in a subset of Eutheria since the last 100 million years.

@@

\subsection{The morphospace}
In this example, we are going to use a subset of the data from \cite{beckancient2014} that contains an ordinated cladistic matrix and a phylogenetic tree containing both the same 50 taxa along side a table that contains some first and last occurence data for several taxa.
The ordinated matrix will represent our full morphospace, i.e. all the morphologies that ever existed through time (for this dataset).

@@

## And a phylogeny:
data(BeckLee_tree)
plot(BeckLee_tree, cex=0.8); axisPhylo(root=140)
@

@@

<<strap, eval=FALSE>>=
if(!require(strap)) install.packages("strap")
geoscalePhylo(BeckLee_tree, cex.tip=0.8, cex.ts=0.6)
@

\section{A disparity-through-time analysis}
\subsection{Splitting the morphospace through time} 
One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller sub-samples that contain the total amount of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
Basically, the full morphospace in the previous step represents the total amount of morphologies in all time and is likely to be greater than any sub-sample of the morphospace.

@@

<<time-binning>>=
## Creating the vector of time bins ages
time_bins <- rev(seq(from=0, to=100, by=20))

@@

The warning message tells us that some taxa were considered as single points in time because we didn't provide temporal range information for all the taxa: they are therefore considered as unique time points without any occurrence range.

The output object is a \dispRity object.
For details about this object class, please refer to the \href{https://github.com/TGuillerme/dispRity}{the \dispRity manual}.
In brief, \dispRity objects are lists of different elements (i.e. disparity results, morphospace sub-samples, etc...) that display only a summary of the object when calling the object for avoiding jamming the \R console.

@@

\subsection{Bootstrapping the data}
Once we obtain our different subsamples, we can bootstrap and rarefy them (i.e. pseudo-replicating the data).
The bootstrapping will allow to make each subsample more robust to outliers and the rarefaction will allow us to compare slices with the same number of taxa to get rid of eventual sampling problems (i.e. more taxa in one bin than the other).
The \texttt{boot.matrix} function that allows to bootstrap the \dispRity object and the \texttt{rarefaction} option within allows to rarefy it.

<<boot.rare>>=
## Bootstrapping each sub-sample 100 times
boot_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100)

## Getting the minimum number of rows (=taxa) in the sub-samples
min(unlist(lapply(binned_morphospace$data, nrow)))

## Bootstrapping each sub-sample 100 times and rarefying them 
rare_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100,
    rarefaction = 6)

@@

We can now calculate the disparity within each sub-sample along side with some confidence intervals generated by the pseudo-replication step above (bootstraps/rarefaction).
Disparity can be calculated in many ways and this package allows user to come up with their own disparity metrics.
For more info, please refer to the \href{https://github.com/TGuillerme/dispRity}{\dispRity manual and the metric vignette}.

In this example, we are going to calculate the spread of the data in each sub-sample by calculating the disparity as the sum of the variance of each dimension of the morphospace in each sub-sample using the \dispRity function.
Thus, in this example, the disparity is defined by the multi-dimensional variance of each sub-sample (i.e. the spread of the taxa within the morphospace).
Note that this metric comes with a caveat not solved here since it ignores the covariance among the dimensions of the morphospace.

@@

## To display the actual results we can use the summary function
summary(boot_disparity)
summary(rare_disparity)

## See more options on the summary function:
?summary.dispRity
@

@@

## Plotting the bootstrapped and rarefied results
plot(boot_disparity, type = "continuous", main = "bootstrapped results")
plot(rare_disparity, type = "continuous", main = "rarefied results")

## See more options on the plot function:
?plot.dispRity
@

@@

\subsection{Testing differences}
Finally, to draw some conclusions from these results, we need to apply some statistical testing.
As stated before, we wanted to test whether the disparity changed through time in this dataset since the last 100 million years.
To do so, we can apply some comparisons of the means between each time-bins in a sequential manner to see whether the disparity in bin X is equal to the disparity in bin Y which is in turn equal to the disparity in bin Z, etc.
Because our data is temporally auto-correlated (i.e. what happens in bin Y depends on what happened in bin X) and was pseudo-replicated (i.e. each bootstrap draw creates non-independent sub-samples because they are all based on the same sub-sample), we are going to apply a non-parametric mean comparison: the \texttt{wilcox.test}.
Also, we are going to apply a p-value correction to avoid type I error inflation due to performing multiple test for the same hypothesis (see \texttt{?p.adjust} for more details).

>>>>

This demo aims to give quick overview of the \dispRity package (v.\Sexpr{packageVersion("dispRity")}) for palaeobiology analysis.
Please refer to \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} for other vignettes, namely the \href{https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.pdf}{\dispRity manual} that explains the functions in more detail.

To keep it short, this package allows to easily perform \textbf{disparity-through-time} analysis.
This type of analysis often starts by ordinating \textbf{morphometric} or \textbf{cladistic} data into a multidimensional object hereafter called the \textbf{morphospace}.
One might be interested in studying how the occupancy of the morphospace changed through time by measuring a summary metric of this morphospace through time, called \textbf{disparity}.
Disparity can then be seen as a value summarising the diversity of morphologies.

This demo showcases a typical disparity-through-time analyse: we are going to test whether the disparity changed through time in a subset of Eutheria (Mammalia) since the last 100 million years.

@@

\subsection{The morphospace}
In this example, we are going to use a subset of the data from \cite{beckancient2014} that contains an ordinated cladistic matrix and a phylogenetic tree containing 50 taxa along side with a table containing some first and last occurence data for several taxa.
The ordinated matrix will represent our full morphospace, i.e. all the morphologies that ever existed through time (for this dataset).

@@

## And a phylogeny:
data(BeckLee_tree)
plot(BeckLee_tree, cex=0.7); axisPhylo(root=140)
@

@@

<<strap, eval=FALSE>>=
if(!require(strap)) install.packages("strap")
library(strap)
geoscalePhylo(BeckLee_tree, cex.tip=0.7, cex.ts=0.6)
@

\section{A disparity-through-time analysis}
\subsection{Splitting the morphospace through time} 
One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller time seriess that contain the total amount of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
Basically, the full morphospace in the previous step represents the total amount of morphologies in all time and is likely to be greater than any time series of the morphospace.

@@

<<time-binning>>=
## Creating the vector of time bins ages
time_bins <- rev(seq(from = 0, to = 100, by = 20))

@@

The output object is a \dispRity object.
For details about this object class, please refer to the \href{https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.pdf}{\dispRity manual}.
In brief, \dispRity objects are lists of different elements (i.e. disparity results, morphospace time seriess, etc...) that display only a summary of the object when calling the object for avoiding jamming the \R console.

@@

\subsection{Bootstrapping the data}
Once we obtain our different time series, we can bootstrap and rarefy them (i.e. pseudo-replicating the data).
The bootstrapping will allow to make each series more robust to outliers and the rarefaction will allow us to compare slices with the same number of taxa to get rid of eventual sampling problems (i.e. more taxa in one bin than the other).
The \texttt{boot.matrix} function that allows to bootstrap the \dispRity object and the \texttt{rarefaction} option within allows to rarefy it.

<<boot.rare>>=
## Bootstrapping each time series 100 times
boot_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100)

## Getting the minimum number of rows (=taxa) in the time seriess
min(unlist(lapply(binned_morphospace$data, nrow)))

## Bootstrapping each time series 100 times and rarefying them 
rare_bin_morphospace <- boot.matrix(binned_morphospace, bootstraps = 100,
    rarefaction = 6)

@@

We can now calculate the disparity within each time series along side with some confidence intervals generated by the pseudo-replication step above (bootstraps/rarefaction).
Disparity can be calculated in many ways and this package allows user to come up with their own disparity metrics.
For more info, please refer to the \href{https://github.com/TGuillerme/dispRity}{\dispRity manual and the metric vignette}.

In this example, we are going to calculate the spread of the data in each time series by calculating the disparity as the sum of the variance of each dimension of the morphospace in each time series using the \dispRity function.
Thus, in this example, the disparity is defined by the multi-dimensional variance of each time series (i.e. the spread of the taxa within the morphospace).
Note that this metric comes with a caveat not solved here since it ignores the covariance among the dimensions of the morphospace.

@@

## To display the actual results we can use the summary function
summary(boot_disparity)
summary(rare_disparity)
@

\noindent As for any \R package, you can refer yourself to each individual function for more details.

<<disparity2, eval = FALSE>>=
## See more options on the summary function:
?summary.dispRity
@

@@

## Plotting the bootstrapped and rarefied results
plot(boot_disparity, type = "continuous", main = "bootstrapped results")
plot(rare_disparity, type = "continuous", main = "rarefied results")
@

@@

\subsection{Testing differences}
Finally, to draw some conclusions from these results, we need to apply some statistical testing.
As stated before, we wanted to test whether the disparity changed through time in this dataset since the last 100 million years.
To do so, we can apply some comparisons of the means between each time-bins in a sequential manner to see whether the disparity in bin X is equal to the disparity in bin Y which is in turn equal to the disparity in bin Z, etc.
Because our data is temporally auto-correlated (i.e. what happens in bin Y depends on what happened in bin X) and was pseudo-replicated (i.e. each bootstrap draw creates non-independent time seriess because they are all based on the same time series), we are going to apply a non-parametric mean comparison: the \texttt{wilcox.test}.
Also, we are going to apply a p-value correction to avoid type I error inflation due to performing multiple test for the same hypothesis (see \texttt{?p.adjust} for more details).
